Leadership across cultures and contexts: an exploratory analysis

### 2020-05-09 leadership_across_cultures_contexts.Rmd
Dissected and rearranged discussion section to match new framework.

Some common leadership dimensions were also relatively invariant across the four factors we examined: group context, continental region, subsistence strategy, and leader sex. These dimensions included qualities, such as being knowledgeable and intelligent, wealthy, strategically nepotistic, and generous, and functions, such as resolving conflict and representing the group. These dimensions are therefore candidates for universal dimensions of human leadership. By "universal" we do not mean that all leaders have all these traits. Instead, we mean that there is evidence for these traits across a solid majority of cultures that does not vary much by the four factors. See Table \@ref(tab:leaderuniversals) and Figure \@ref(fig:varsupport).

@@

Leaders can be conceptualized as performing three broad functions: providing goods and services (Prosocial), strategizing, and organizing collective actions.

The leadership "features" we derived by clustering our leadership dimensions were represented across most cultures. See Figure \@ref(fig:featuressupportplot).

@@

The universal dimensions also partially correspond to the service-for-prestige model [@price_evolution_2014; @price_service-for-prestige_2015], but that model does not emphasize the concrete material benefits seen here. The absence of *High status* and *Aggressiveness* on this universal list corresponds to the reverse dominance hierarchy model [@boehm_egalitarian_1993], which emphasizes knowledge, intelligence, conflict resolution, generosity and status leveling mechanisms.

@@

*Coercive authority* is one leadership dimension that is common in the ethnographic record and for which evidence does not vary by any of our four factors, yet did not quite make the 60% cutoff to be included on our list of candidate universals. This dimension features in several evolutionary models of leadership involving dominance hierarchies [@barkow_darwin_1989; @tiger_imperial_1971; @henrich_evolution_2001] and the role of punishment in promoting cooperation [e.g., @marlowe_more_2008; @henrich_markets_2010; @henrich_costly_2006; @boyd_punishment_1992]. Indeed, we found that this dimension was associated with *Punishment*, *Aggressiveness*, and *Feared*, and also with *Control economics*, which evokes "resource holding power", a concept from the behavioral ecology of dominance hierarchies. See Figure \@ref(fig:elasticcoercive). This passage describing Trobriand chiefs is illustrative [@oj29-007, pp. 64-65]:

@@

There was substantial cross-cultural variation in evidence for most dimensions of leadership: of our `r length(all_study_vars)` dimensions of leadership, `r sum(df_compare2$Estimate < 0.5)` had evidence in less than 50% of our `r nrow(leader_cult)` cultures. See Figures \@ref(fig:varsupport) and \@ref(fig:varsupportcb). There was *systematic* variation in `r vary_n` leadership dimensions by continental region, subsistence, strategy, group context, and/or leader sex, factors for which we had complete data for all text records. Interestingly, these systematically varying dimensions tended to be associated with one factor above the others. See Figure \@ref(fig:heatmapdrop1). The clearest pattern to emerge from these analyses is that, not too surprisingly, many dimensions of leadership are context-specfic: Aggressiveness and Bravery were associated with military groups, Supernatural and Ritual functions were associated with religious groups, Organizing cooperation and Experience/accomplished were associated with economic groups, and Providing counsel and older Age were associated with kin groups. See Figure \@ref(fig:heatmapgroups).

Evidence for many leadership dimensions did vary across cultures and contexts, however, including functions such as organizing cooperation, providing counsel and direction, and prosocial investment; and qualities such as high status and aggressiveness. See Figure \@ref(fig:heatmapdrop1).

Most discussions of leadership in the ethnographic record pertained to male leaders of community-level political groups, with leadership in residential subgroups and kin groups relatively more common in hunter-gatherer and pastoralist societies, and supracommunity political groups more common in horticultural and agricultural societies. Leadership in all social contexts, including economic, military, and religious, was present across all subsistence strategies, however (see Figure \@ref(fig:grouptype)). The relative unimportance of supracommunity leadership in hunter-gatherers and pastoralists, and the relative importance of leadership in residential subgroups, highlights a trend within non-WEIRD populations in sociopolitical organizations, as these generally more sedentary populations with more intensified subsistence strategies more closely resemble sociopolitical organization of WEIRD societies. Thus, these results provide a fine-grained view of cultural transitions in leadership systems.

## Subsistence strategy and group context effects on human leadership

The distribution of text records by group and subsistence type suggests leaders emerge within similar social contexts among populations with varied subsistence strategies, but at variable frequencies. See Figure \@ref(fig:grouptype). Exploratory analyses identified both universal and variable dimensions of human leadership across cultures and contexts.

### Universal dimensions

Exploratory modeling revealed variation in the vast majority (INCLUDE NO. or %) of our researcher coded variables (TABLEs) did not vary by group type or subsistence strategy (adjusting for effects of leader sex and continental region and the hierarchical structure of ethnographer within culture), suggesting that many of these dimensions of leadership are potentially universal features of human social groups or at least not strongly impacted by the within or between-group measures investigated here.

### Variable dimensions

Leadership in residential subgroups (e.g., social groups, performance groups) was relatively common among hunter-gatherers but relatively rare among agriculturists. Leadership at the supracommunity level, in contrast, was relatively rare among hunter-gatherers but quite common among agriculturists. The only example of supracommunity leadership among hunter-gatherers was from the Andamanese, a group of populations of Negrito hunter-gatherers of the Andaman Islands by [@az02-001, p. 45], who discussed the roles of a "head chief" from a permanent residential location and their marginal influence over other, local residential leaders (see quote in Methods section).

Hunter-gatherers and pastoralists, populations more likely to experience high residential mobility [@wild_making_2019; @rigby_persistent_1985; @kelly_foraging_2013], exhibited greater relative frequencies of leadership within residential groups and kin groups. Pastoralists, however, rely in part on a highly defensible resource base (i.e., livestock) and exhibit leadership beyond the community more so than hunter-gatherers and on par with horticulturalists. Pastoral and horticultural populations often experience elevated rates of inter-group conflict relative to mobile hunter-gatherers and these results are consistent with a relationship between increased inter-group conflict and more elaborated political leadership systems [@ember_warfare_1992]. For example, among the Somali, a politically complex and socially stratified agro-pastoralists group distributed across the Horn of Africa, @mo04-021 [p. 58] notes that, "the Sultan does not lead the army of the tribe in war. He names a chief of the armed forces."

## Leader functions and qualities

Our analysis identified several leader functions and qualities that were near cultural universals, and others that were more variable or relatively rare. See Figure \@ref(fig:varsupport).

<!-- NEED TO DO ADD SECTION OR DELETE HERE -->

### Variable dimensions

@@

@@

Several theories of leadership also emphasize the moral authority of leaders [e.g., @henrich_big_2015; @henrich_evolution_2001; @richerson_tribal_2009; @tietjen_moral_1985; @boggild_evolved_2016], yet leaders were seen as sources of moral authority or "fair" in only a minority of cultures (`r moral_pct`% and `r fairness_pct`% of cultures, respectively). The *Dominance* theory of leadership [e.g., @chapais_competence_2015; @cheng_dominance_2019; @cheng_pride_2010; @henrich_evolution_2001; @tiger_imperial_1971] emphasizes leaders' ability to instill fear, but in only `r feared_pct`% of cultures were leaders characterized as feared. Finally, a large literature on leadership in industrialized societies highlights the physical attractiveness of leaders [e.g., @sinclair_sexuality_1995; @altemeyer_sexual_1974; @march_leadership_2009], but leaders were characterized as attractive in only `r attractive_pct`% of cultures.

WARFARE H&G Figure S\@ref(fig:heatmapsubsis).

### Disentangling the relationship between leadership and high status

@@

## Features of human leadership

We performed hierarchical cluster analysis on our leader quality and leader function variables to determine if these variables formed theoretically meaningful groups. Variables were clustered together if they were positively correlated. Leadership is expected to be associated with particular suites of characteristics and functions, which potentially vary by cultural or social context, personality, or via other inter-individual variation [@cheng_two_2013; @von_rueden_roots_2014; @hogan_personality_2020; @lukaszewski_extraversion_2015; @henrich_evolution_2001; @van_vugt_dual_2019]. Much of the evolutionary theoretical literature on leadership has suggested social influence is most often achieved via one of two distinct strategies: prestige or dominance. @van_vugt_dual_2019 [p. 1], for example, "argue that there are two leadership styles: prestige style and dominance style," and @cheng_two_2013 [p. 119] in focusing on dominance and prestige as two distinct leadership strategies suggest that, "human social stratification is characterized by the co-occurrence of both strategies."

The dominance-prestige theory as formulated by @henrich_evolution_2001 significantly advanced evolutionary approaches to leadership and inspired much subsequent theoretical and empirical work. Exploratory cluster analyses reported here, however, suggest multiple features of leader qualities and leader functions across cultures and contexts, beyond dominance and prestige. Leaders likely rely a range of individual capacities, including physical, supernatural, and cognitive endowments, in order to organize group members, mediate problems, and provide prosocial services to the group, all while conforming to social and cultural pressures.

### Clusters of leader qualities

Leader qualities formed two top-level clusters, which we identify as a "Prosocial qualities" cluster, and an "Individual qualities" cluster, each of which contained two interpretable and generally supported subclusters. See Figure \@ref(fig:cluster)A.

### Clusters of leader functions

@@

## Limitations

@@

>>>>

<!-- Most discussions of leadership in the ethnographic record pertained to male leaders of community-level political groups...

Leadership in all social contexts, including economic, military, and religious, was present across all subsistence strategies, however (see Figure \@ref(fig:grouptype)).
-->

Some common leadership dimensions were also relatively invariant across the four factors we examined: group context, subsistence strategy, continental region, and leader sex. These dimensions included qualities, such as being knowledgeable and intelligent, wealthy, strategically nepotistic, and generous, and functions, such as resolving conflict and representing the group. These dimensions are therefore candidates for universal dimensions of human leadership. By "universal" we do not mean that all leaders have all these traits. Instead, we mean that there is evidence for these traits across a solid majority of cultures that does not vary much by the four factors. See Table \@ref(tab:leaderuniversals) and Figure \@ref(fig:varsupport).

@@

Leaders can be conceptualized as performing three broad functions: providing prosocial goods and services, strategizing, and organizing collective actions. Regarding the traits charachteristic of leadership, leaders can be conceptualized as possessing some combination of individual capacities, many of which are necessarily prosocial, conforming to particular cultural pressures, and recieving various benefits in exchange for their functions. The leadership "features" we derived by clustering leadership dimensions were represented across most cultures. See Figure \@ref(fig:featuressupportplot).

@@

The universal dimensions also partially correspond to the service-for-prestige model [@price_evolution_2014; @price_service-for-prestige_2015] in which leaders recieve special status in exchange for the services they provide followers, but that model does not emphasize the concrete material benefits seen here. The absence of *High status* and *Aggressiveness* on this universal list corresponds to the reverse dominance hierarchy model [@boehm_egalitarian_1993], which emphasizes knowledge, intelligence, conflict resolution, generosity and status leveling mechanisms.

@@

Coercive authority is a dimension of leadership that is common in the ethnographic record and for which evidence does not vary by any of our four factors, yet did not quite make the 60% cutoff to be included on our list of candidate universals. This dimension features in several evolutionary models of leadership involving dominance hierarchies [@barkow_darwin_1989; @tiger_imperial_1971; @henrich_evolution_2001; @cheng_dominance_2019; @van_vugt_dual_2019] and the role of punishment in promoting cooperation [e.g., @marlowe_more_2008; @henrich_markets_2010; @henrich_costly_2006; @boyd_punishment_1992]. Indeed, we found that coercive authoiry among leaders was associated with enforcing punishment, aggressiveness, being feared, and also with controling economics, which evokes "resource holding power", a concept from the behavioral ecology of dominance hierarchies. See Figure \@ref(fig:elasticcoercive). This passage describing Trobriand chiefs is illustrative [@oj29-007, pp. 64-65]:

@@

@@

There was substantial cross-cultural variation in evidence for most dimensions of leadership: of our `r length(all_study_vars)` dimensions of leadership, `r sum(df_compare2$Estimate < 0.5)` had evidence in less than 50% of our `r nrow(leader_cult)` cultures. See Figures \@ref(fig:varsupport) and \@ref(fig:varsupportcb). There was *systematic* variation in `r vary_n` leadership dimensions by group context, subsistence strategy, continental region, and/or leader sex, factors for which we had complete data for all text records. Interestingly, these systematically varying dimensions tended to be associated with one factor above the others. See Figure \@ref(fig:heatmapdrop1). The clearest pattern to emerge from these analyses is that, not too surprisingly, many dimensions of leadership are context-specfic: aggressiveness and bravery were associated with military leadership, supernatural qualities and ritual functions were associated with religious leadership, organizing cooperation and being experience or accomplished were associated with economic-group ledaership, and providing counsel and older age were associated with kin-group leadership. See Figure \@ref(fig:heatmapgroups).

The distribution of text records by group context and subsistence strategy suggests leaders emerge within similar social contexts across populations, but at variable frequencies. See Figure \@ref(fig:grouptype). Leadership in residential subgroups and kin groups was relatively more common in hunter-gatherer and pastoralist societies, populations more likely to experience high residential mobility [@wild_making_2019; @rigby_persistent_1985; @kelly_foraging_2013]. Supracommunity political leadership was more common in horticultural and agricultural societies, which tend to rely heavily on cultivated land and maintain fixed, defensible territories. These patterns support a relationship between residential mobility and subsistence intensification with sociopolitical organization among non-WEIRD populations. The only example of supracommunity leadership among hunter-gatherers was from the Andamanese hunter-gatherers of the Andaman Islands by [@az02-001, p. 45], who discussed the roles of a "head chief" from a permanent residential location and their marginal influence over other, local residential leaders (see quote in Methods section). 
<!-- Thus, these results provide a fine-grained empirical view of putative cultural transitions in leadership systems. -->

Pastoralists, however, rely in part on a highly defensible resource base (i.e., livestock) and exhibit leadership beyond the community more so than hunter-gatherers and on par with horticulturalists. Pastoral and horticultural populations often experience elevated rates of inter-group conflict relative to mobile hunter-gatherers and these results are consistent with a relationship between increased inter-group conflict and more elaborated political leadership systems [@ember_warfare_1992]. For example, among the Somali, a politically complex and socially stratified agro-pastoralists group distributed across the Horn of Africa, @mo04-021 [p. 58] notes that, "the Sultan does not lead the army of the tribe in war. He names a chief of the armed forces."

@@

Several theories of leadership also emphasize the moral authority of leaders [e.g., @henrich_big_2015; @henrich_evolution_2001; @richerson_tribal_2009; @tietjen_moral_1985; @boggild_evolved_2016], yet leaders were seen as sources of moral authority or "fair" in only a minority of cultures (`r moral_pct`% and `r fairness_pct`% of cultures, respectively). The *Dominance* theory of leadership [e.g., @chapais_competence_2015; @cheng_dominance_2019; @cheng_pride_2010; @henrich_evolution_2001; @tiger_imperial_1971] emphasizes leaders' ability to instill fear, but in only `r feared_pct`% of cultures were leaders characterized as feared. Finally, a large literature on leadership in industrialized societies highlights the physical attractiveness of leaders [e.g., @sinclair_sexuality_1995; @altemeyer_sexual_1974; @march_leadership_2009], but leaders were characterized as attractive in only `r attractive_pct`% of the nonindustiral cultures sampled here.

<!-- WARFARE H&G Figure S\@ref(fig:heatmapsubsis).  -->

## Disentangling the relationship between leadership and high status

@@

@@

## Beyond dual models of leadership

Leadership is expected to be associated with particular suites of characteristics and functions, which potentially vary by cultural or social context, personality, or via other inter-individual variation [@cheng_two_2013; @von_rueden_roots_2014; @hogan_personality_2020; @lukaszewski_extraversion_2015; @henrich_evolution_2001; @van_vugt_dual_2019], and as discussed, systematic cross-cultural evidence here supports many aspects of influential theories. Much of the evolutionary theoretical literature on leadership, however, has suggested social influence is most often achieved via one of two distinct strategies: prestige or dominance. The dominance-prestige theory as formulated by @henrich_evolution_2001 significantly advanced evolutionary approaches to leadership and inspired much subsequent theoretical and empirical work. @van_vugt_dual_2019 [p. 1], for example, "argue that there are two leadership styles: prestige style and dominance style," and @cheng_two_2013 [p. 119] in focusing on dominance and prestige as two distinct leadership strategies suggest that, "human social stratification is characterized by the co-occurrence of both strategies." Exploratory cluster analyses reported here, however, suggest multiple features of leader qualities and leader functions across cultures and contexts, beyond dominance and prestige. Leaders likely rely a range of individual capacities, including physical, supernatural, and cognitive endowments, in order to organize group members, impliment strategic actions, and provide prosocial services to the group, all while conforming to social and cultural pressures.

# Limitations

@@

asf
